---
title: "FBI Crime Data"
description: |
  Analysis of FBI crime data API from TidyTuesday
author: Rhyan Goozh
date: Febuary 18, 2025
format: html
execute: 
  warning: false
  message: false
---

In this project, I analyze agency level crime data across all 50 states using the Tidy Tuesday *Agencies From the FBI Crime Data* *API* data set. The data set provides detailed information on law enforcement agencies across all 50 states that have submitted data to the FBI's Uniform Crime Reporting (UCR) program and are displayed on the Crime Data Explorer (CDE). This data set is ideal for exploring trends and understanding the fluidity in crime.

```{r}
#install.packages("tidytuesdayR")
tuesdata <- tidytuesdayR::tt_load('2025-02-18')
agencies <- tuesdata$agencies
```

```{r}
library(tidyverse)
library(ggplot2)
library(tidytuesdayR)

agencies |>
  count(state, agency_type) |>
  ggplot(aes(x = state, y = n, color = agency_type)) +
  geom_point(size = 3) +
  labs(
    x = "State",
    y = "Number of Agencies",
    title = "Distribution of Law Enforcement Agencies by State and Type",
    color = "Agency Type"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "bottom"
    )
```

Here I created a scatterplot that visualizes the distribution of law enforcement agencies by state and agency type. The x-axis represents the states, while the y-axis displays the number of agencies in each state. Different colors represent different agency types.

This visualization helps identify variations in agency distribution across states and highlights trends in the presence of different law enforcement agency types. It can be useful for understanding the allocation of law enforcement resources and regional differences in agency structures.

**Data Source:**

<https://github.com/rfordatascience/tidytuesday/blob/master/data/2025/2025-02-18/agencies.csv>

**Author:** TidyTuesday Project

This data was sourced from the TidyTuesday project, which provides weekly data challenges for the R community. Specifically, the dataset focuses on law enforcement agencies across the U.S.:

\- The dataset was last committed on February 18, 2025.

\- It includes details on agencies that have submitted data to the FBI’s Uniform Crime Reporting (UCR) Program.

\- The dataset provides insights into agency types, geographical distribution, and NIBRS participation.
