---
title: "Permutation Tests"
description: |
  Permutation Test on NHL Player Birthdates vs. Canadian Population
author: Rhyan Goozh
date: March 30, 2025
format: html
execute: 
  warning: false
  message: false
---

This analysis investigates whether NHL players are disproportionately born in certain months compared to the general Canadian population. Specifically, we examine if players are more likely to be born in the first quarter of the year (January–March), potentially due to age-based advantages in youth hockey systems.

Our research question is: Are NHL players more likely to be born in the first quarter of the year than we would expect if their birth months followed the same distribution as the general Canadian population?

Hypotheses:

-   Null hypothesis (H0): The probability that an NHL player is born in the first quarter of the year is equal to the probability that a Canadian is born in the first quarter of the year.

-   Alternative hypothesis (H1): The probability that an NHL player is born in the first quarter of the year is greater than the probability that a Canadian is born in the first quarter of the year.

```{r}
library(tidytuesdayR)

# Load the data for January 9, 2024
tuesdata <- tidytuesdayR::tt_load('2024-01-09')
```

```{r}
library(dplyr)
library(lubridate)
library(readr)
library(tidyverse)
library(ggplot2)
library(purrr)
library(tidyr)
set.seed(123)
```

```{r}
# Importing datasets I will analyze 
canada_births_1991_2022 <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-01-09/canada_births_1991_2022.csv')
nhl_player_births <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-01-09/nhl_player_births.csv')
nhl_rosters <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-01-09/nhl_rosters.csv')
nhl_teams <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-01-09/nhl_teams.csv')
```

```{r}
# Extracting the birth month and filtering relevant data from NHL dataset
nhl_player_births <- nhl_player_births |> 
  mutate(birth_month = month(as.Date(birth_date, format = "%Y-%m-%d"))) |> 
  filter(!is.na(birth_month))

# Preprocessing Canadian data for birth months in the general population and removing NA values
canada_births_1991_2022 <- canada_births_1991_2022 |> 
  filter(!is.na(month))
```

```{r}
# Summary Statistics
# Plot visualization of the distributions of birth months for NHL players and the Canadian population

# Plot for NHL players
ggplot(nhl_player_births, aes(x = birth_month)) +  # Use 'birth_month' from nhl_player_births
  geom_bar(fill = "skyblue", color = "black") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(title = "Birth Month Distribution of NHL Players", x = "Month", y = "Count") +
  theme_minimal()

# Plot for Canadian population
ggplot(canada_births_1991_2022, aes(x = month)) +  # Use 'month' from canada_births_1991_2022
  geom_bar(fill = "orange", color = "black") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(title = "Birth Month Distribution of Canadian Population", x = "Month", y = "Count") +
  theme_minimal()
```

```{r}
# Summary Statistics
# Calculating the proportion of players and population born in the first quarter (Jan-Mar)

# NHL first quarter proportion
nhl_first_quarter <- nhl_player_births |> 
  filter(birth_month %in% c(1, 2, 3)) |> 
  summarize(first_quarter_prop = n() / nrow(nhl_player_births))

# Canadian first quarter proportion
canadian_first_quarter <- canada_births_1991_2022 |> 
  filter(month %in% c(1, 2, 3)) |>  # 'month' column in the Canada dataset
  summarize(first_quarter_prop = n() / nrow(canada_births_1991_2022))

# View results
nhl_first_quarter
canadian_first_quarter
```

```{r}
perm_test <- function(nhl_data, canadian_data, n_perm = 5000) {
  # Combine both datasets into one
  all_months <- c(nhl_data$birth_month, canadian_data$month)
  group_labels <- c(rep("NHL", nrow(nhl_data)), rep("Canada", nrow(canadian_data)))
  
  # Observed difference
  observed_diff <- mean(nhl_data$birth_month %in% 1:3) - mean(canadian_data$month %in% 1:3)
  
  # Permutation distribution
  perm_diffs <- replicate(n_perm, {
    shuffled_labels <- sample(group_labels)
    q1_prop_nhl <- mean(all_months[shuffled_labels == "NHL"] %in% 1:3)
    q1_prop_canada <- mean(all_months[shuffled_labels == "Canada"] %in% 1:3)
    q1_prop_nhl - q1_prop_canada
  })
  
  tibble(observed = observed_diff, permuted = perm_diffs)
}
```

We conduct a right-tailed permutation test to evaluate whether a higher proportion of NHL players are born in the first quarter of the year (January–March) compared to the general Canadian population. If the resulting p-value is below the significance level (0.05), we reject the null hypothesis that birth month distributions are the same. Otherwise, we fail to reject it.
```{r}
# Run the permutation test
perm_results <- perm_test(nhl_player_births, canada_births_1991_2022)

# Extract observed and permuted differences
observed_diff <- perm_results$observed[1]
permuted_diffs <- perm_results$permuted

# Calculate p-value
p_value <- mean(permuted_diffs >= observed_diff)
p_value
```
The permutation test shows that the proportion of NHL players born in Q1 (January–March) is significantly higher than expected compared to the Canadian population. With a p-value of 0.0232, we reject the null hypothesis that the birth month distributions are the same, suggesting a real bias in birth month representation among NHL players.

Conclusion:

The permutation test indicates that the proportion of NHL players born in the first quarter of the year (January–March) is significantly higher than what we would expect based on the distribution of birth months in the general Canadian population. With a p-value of 0.0232, we reject the null hypothesis that the birth month distributions are the same. This suggests a real bias in birth month representation among NHL players, where they are more likely to be born in the first quarter of the year.

While this finding aligns with the concept of the relative age effect (RAE)—where players born earlier in the year may have a competitive advantage—it's important to recognize that other factors such as sample size, dataset variations, or specific characteristics of the NHL player pool could influence these results. Further research might consider these factors to deepen our understanding of birth month biases in hockey and other sports.

**Data Source:**

This analysis uses data from the TidyTuesday project for the week of January 9, 2024, accessed via the tidytuesdayR package and the TidyTuesday GitHub repository: 
https://github.com/rfordatascience/tidytuesday/tree/main/data/2024/2024-01-09 

Direct provenance for each dataset:

Canadian Births (1991-2022): This dataset includes monthly birth counts for Canadian residents from 1991 to 2022. It is available at: https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710000901

NHL Player Births, NHL Rosters, NHL Teams: This dataset contains birthdates of NHL players. The original data can be accessed through the NHL API:
https://www.nhl.com/stats/ 

All datasets are publicly available via GitHub and were accessed using the tidytuesdayR package.
