---
title: "SQL"
description: | 
  Traffic Stop Outcomes by Race and Gender Across Three States
author: Rhyan Goozh
date: April 24, 2025
format: html
code-fold: true
execute: 
  warning: false
  message: false
---

```{r}
library(RMariaDB)

con_traffic <- DBI::dbConnect(
 RMariaDB::MariaDB(),
 dbname = "traffic",
 host = Sys.getenv("TRAFFIC_HOST"),
 user = Sys.getenv("TRAFFIC_USER"),
 password = Sys.getenv("TRAFFIC_PWD")
)
```

## Introduction

This project explores how traffic stop outcomes vary by race and gender across California, Texas, and North Carolina. Using the Stanford Open Policing Project SQL database, we query three state-wide traffic stop datasets to identify potential disparities in stop outcomes such as arrests, searches, or warnings. Our analysis includes grouped summaries and visualizations to highlight trends and differences across the states.

## Table Previews

```{sql, connection=con_traffic}
SHOW TABLES;
```

```{sql, connection=con_traffic}
SELECT * FROM ca_statewide_2023_01_26 LIMIT 5;
```

```{sql, connection=con_traffic}
SELECT * FROM tx_statewide_2020_04_01 LIMIT 5;
```

```{sql, connection=con_traffic}
SELECT * FROM nc_statewide_2020_04_01 LIMIT 5;
```

## Checking for Missing Values

```{sql, connection=con_traffic}
-- Check missing subject_race in CA
SELECT COUNT(*) FROM ca_statewide_2023_01_26 WHERE subject_race IS NULL;
```
```{sql, connection=con_traffic}
-- Check missing subject_sex in CA
SELECT COUNT(*) FROM ca_statewide_2023_01_26 WHERE subject_sex IS NULL;
```
```{sql, connection=con_traffic}
-- Check missing outcome in CA
SELECT COUNT(*) FROM ca_statewide_2023_01_26 WHERE outcome IS NULL;
```
California Missing Values:
There is 1 missing value for subject_race, suggesting almost complete data in this category. The subject_sex field has 96 missing values, which is still minimal relative to the dataset size. However, the outcome variable is missing in 9,588,775 records, indicating a severe gap in outcome reporting that could significantly limit analysis and raise questions about data collection methods or system design.

```{sql, connection=con_traffic}
-- Check missing subject_race in TX
SELECT COUNT(*) FROM tx_statewide_2020_04_01 WHERE subject_race IS NULL;
```
```{sql, connection=con_traffic}
-- Check missing subject_sex in TX
SELECT COUNT(*) FROM tx_statewide_2020_04_01 WHERE subject_sex IS NULL;
```
```{sql, connection=con_traffic}
-- Check missing outcome in TX
SELECT COUNT(*) FROM tx_statewide_2020_04_01 WHERE outcome IS NULL;
```
Texas Missing Values:
The dataset shows 352 missing values for subject_race and 436 missing values for subject_sex, which are relatively small proportions and suggest generally good data quality for these variables. There are 2,357 missing values for outcome, a non-negligible amount that should be noted in any evaluation of the dataset’s completeness.

```{sql, connection=con_traffic}
-- Check missing subject_race in NC
SELECT COUNT(*) FROM nc_statewide_2020_04_01 WHERE subject_race IS NULL;
```
```{sql, connection=con_traffic}
-- Check missing subject_sex in NC
SELECT COUNT(*) FROM nc_statewide_2020_04_01 WHERE subject_sex IS NULL;
```
```{sql, connection=con_traffic}
-- Check missing outcome in NC
SELECT COUNT(*) FROM nc_statewide_2020_04_01 WHERE outcome IS NULL;
```
North Carolina Missing Values:
Both subject_race and subject_sex have 18 missing values each, reflecting strong consistency in demographic reporting. On the other hand, 605,059 records are missing outcome values, which could meaningfully affect any modeling or statistical inference relying on outcomes and highlights the importance of addressing or accounting for this missingness.

## SQL Table Queries

```{sql, connection=con_traffic}
-- Query 1: CA outcomes by race and gender
SELECT 
  'CA' AS state,
  subject_race,
  subject_sex,
  outcome,
  COUNT(*) AS count
FROM ca_statewide_2023_01_26
WHERE subject_race IS NOT NULL 
  AND subject_sex IS NOT NULL 
  AND outcome IS NOT NULL
GROUP BY subject_race, subject_sex, outcome
ORDER BY count DESC
```

```{sql, connection=con_traffic}
-- Query 2: TX outcomes by race and gender
SELECT 
  'TX' AS state,
  subject_race,
  subject_sex,
  outcome,
  COUNT(*) AS count
FROM tx_statewide_2020_04_01
WHERE subject_race IS NOT NULL 
  AND subject_sex IS NOT NULL 
  AND outcome IS NOT NULL
GROUP BY subject_race, subject_sex, outcome
ORDER BY count DESC
```

```{sql, connection=con_traffic}
-- Query 3: NC outcomes by race and gender
SELECT 
  'NC' AS state,
  subject_race,
  subject_sex,
  outcome,
  COUNT(*) AS count
FROM nc_statewide_2020_04_01
WHERE subject_race IS NOT NULL 
  AND subject_sex IS NOT NULL 
  AND outcome IS NOT NULL
GROUP BY subject_race, subject_sex, outcome
ORDER BY count DESC
```

## Load Data into R from SQL Queries

```{r}
# Query CA results
ca_data <- DBI::dbGetQuery(con_traffic, "
  SELECT 
    'CA' AS state,
    subject_race,
    subject_sex,
    outcome,
    COUNT(*) AS count
  FROM ca_statewide_2023_01_26
  WHERE subject_race IS NOT NULL 
    AND subject_sex IS NOT NULL 
    AND outcome IS NOT NULL
  GROUP BY subject_race, subject_sex, outcome
  ORDER BY count DESC
")

# Query TX results
tx_data <- DBI::dbGetQuery(con_traffic, "
  SELECT 
    'TX' AS state,
    subject_race,
    subject_sex,
    outcome,
    COUNT(*) AS count
  FROM tx_statewide_2020_04_01
  WHERE subject_race IS NOT NULL 
    AND subject_sex IS NOT NULL 
    AND outcome IS NOT NULL
  GROUP BY subject_race, subject_sex, outcome
  ORDER BY count DESC
")

# Query NC results
nc_data <- DBI::dbGetQuery(con_traffic, "
  SELECT 
    'NC' AS state,
    subject_race,
    subject_sex,
    outcome,
    COUNT(*) AS count
  FROM nc_statewide_2020_04_01
  WHERE subject_race IS NOT NULL 
    AND subject_sex IS NOT NULL 
    AND outcome IS NOT NULL
  GROUP BY subject_race, subject_sex, outcome
  ORDER BY count DESC
")
```

## Combine States Data

```{r}
# Combine the data from CA, TX, and NC
combined_data <- rbind(ca_data, tx_data, nc_data)

# Convert 'state' to a factor for easier plotting
combined_data$state <- factor(combined_data$state, levels = c('CA', 'TX', 'NC'))

# Convert 'subject_sex' and 'subject_race' to factors
combined_data$subject_sex <- factor(combined_data$subject_sex)
combined_data$subject_race <- factor(combined_data$subject_race)
```

## Two Well Labled Plots

```{r}
# First Plot: Traffic stop outcomes by race across the states
library(dplyr)
library(ggplot2)
library(scales)

all_data <- bind_rows(ca_data, tx_data, nc_data)

race_summary <- all_data |>
  group_by(state, subject_race) |>
  summarise(total_stops = sum(count), .groups = 'drop')

ggplot(race_summary, aes(x = subject_race, y = total_stops, fill = state)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +  # Added this to format the y-axis
  labs(
    title = "Traffic Stop Outcomes by Race Across States",
    x = "Race",
    y = "Number of Traffic Stops",
    fill = "State",
    caption = "Data from the Stanford Open Policing Project"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
The bar chart compares the number of traffic stops by race across California, North Carolina, and Texas. Across all three states, White drivers experienced the highest number of stops, followed by Hispanic and Black drivers. There is a significant drop-off in the number of stops for drivers categorized as Asian/Pacific Islander, "Other," and "Unknown," indicating that stops are heavily concentrated among a few racial groups. Texas is leading in the number of stops for White, Hispanic, and Unknown drivers. North Carolina shows a particularly high number of stops for Black drivers compared to California and Texas.

```{r}
# Second Plot: Traffic stop outcomes by gender across the states
library(scales)

gender_summary <- all_data |>
  group_by(state, subject_sex) |>
  summarise(total_stops = sum(count), .groups = 'drop')

ggplot(gender_summary, aes(x = subject_sex, y = total_stops, fill = state)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +  # Added this to format the y-axis
  labs(
    title = "Traffic Stop Outcomes by Gender Across States",
    x = "Gender",
    y = "Number of Traffic Stops",
    fill = "State",
    caption = "Data from the Stanford Open Policing Project"
  ) +
  theme_minimal()
```
This plot shows traffic stop outcomes by gender across California, North Carolina, and Texas. Male drivers account for a noticeably higher number of stops in each state compared to female drivers. The gap between male and female stops is consistent, but Texas records a particularly large number of male stops relative to the other two states. California has about 5 million more male stops than North Caroline but the number of stop for women between the two states is relatively the same. Overall, gender appears to be a significant factor in the frequency of traffic stops.

## Insights

This analysis shows how the frequency and type of traffic stop outcomes (e.g., warnings, arrests, searches) vary by race, gender, and state. Across California, Texas, and North Carolina, white male drivers are stopped most often, but Black and Hispanic individuals—especially males—face higher rates of searches and arrests, particularly in Texas and North Carolina. White drivers, especially females, are more likely to receive warnings. These trends point to consistent disparities and potential bias in traffic enforcement across state lines.

## Reference

Pierson, Emma, Camelia Simoiu, Jan Overgoor, Sam Corbett-Davies, Daniel Jenson, Amy Shoemaker, Vignesh Ramachandran, et al. 2020. “A Large-Scale Analysis of Racial Disparities in Police Stops Across the United States.” Nature Human Behaviour, 1–10.
